Mapping mutational effects along the evolutionary landscape of HIV envelope

### 2018-02-25 paper/manuscript.tex
Add supp fig on clusters of shifted sites

Specifically, we used our deep mutational scanning data to construct experimentally informed codon models (ExpCM's) for Env's evolution.
An ExpCM is a phylogenetic substitution model that incorporates the functional constraints measured in a deep mutational scanning experiment~\citep{hilton2017phydms}.
If the experiment captures much of the actual evolutionary constraint on a gene, then an ExpCM will describe the gene's natural evolution better than a standard phylogenetic codon substitution model.
The reason is that standard codon substitution models~\citep[such as the commonly used Goldman-Yang style models;][]{yang2000codon} only model functional constraint via a single parameter that represents the rate of fixation of nonsynonymous protein-altering mutations relative to synonymous ones; this parameter is called dN/dS or $\omega$.
In contrast, an ExpCM accounts for the preference of each site for each of the 20 amino acids under the functional selection in the deep mutational scan, and then additionally adds an $\omega$ parameter that represents the relative rate of nonsynonymous to synonymous substitutions after accounting for these functional constraints~\citep{bloom2017identification,hilton2017phydms}.
Importantly, since we expect some sites in Env to be under diversifying selection from immunity, we extended the ExpCM's described in \citet{hilton2017phydms} to draw $\omega$ from a gamma distribution as is commonly done for codon-substitution models~\citep{yang2000codon}.

@@

\begin{figure}
\begin{fullwidth}
{\includegraphics[width=1.3\textwidth]{figures/BG505_prefs.pdf}}
\caption{\label{fig:BG505prefs}
Amino-acid preferences for the BG505 Env.
At each site, the height of the letter is proportional for that site's preference for that amino acid.
The top color bar indicates the region of Env (gp120 variable loop, gp120 not variable loop, or gp41).
The lower color bar indicates the evidence that the site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experiments~\citep[see the last section of the Results and][]{bloom2017identification}.
We report the $P$-value for $\omega_r \ne 1$ rather than the value of $\omega_r$ itself since point estimates of $\omega_r$ are unreliable for individual sites due to low numbers of observations, making the $P$-value a better indicator of the strength of the statistical evidence for faster or slower than expected evolution~\citep{kosakovsky2005not,murrell2012detecting}.
The letters above the logos indicate the wildtype amino acid in BG505.
Sites are numbered using the HXB2 scheme~\citep{korber1998numbering}.
This logo plot shows the site-specific amino-acid preferences for BG505 after averaging the replicates and re-scaling by the stringency parameter in \TABLE{phydms}.
The figure was generated using \texttt{dms\_tools2}~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}, which in turn utilizes \texttt{weblogo}~\citep{crooks2004weblogo}.
}
\figdata{The numerical values of the amino-acid preferences plotted in this figure are in \texttt{rescaled\_BG505\_prefs.csv.}}
\figdata{The sequence of BG505 Env and mapping from sequential (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BG505\_to\_HXB2.csv}.}
\figdata{The $\omega_r$ values and associated $P$-values for BG505 in HXB2 numbering are in \texttt{BG505\_omegabysite.tsv}.}
\end{fullwidth}
\end{figure}

@@

\figsupp[\label{figsupp:shifts_on_six_helix_bundlel}Cluster of shifted and substituted sites in context of the post-fusion six-helix bundle of Env.]
{This figure shows the six-helix bundle of Env~\citep[PDB 1ENV;][]{weissenhorn1997atomic}, focusing on the central three helices, each of which is derived from a different Env monomer and which pack together to form the bundle's core.
The $\sim$80 residues per monomer that are resolved in this structure include four significantly shifted sites and five substituted sites.
Of these, three of the shifted sites (582, 583, and 587) and one substituted site (588) form a cluster at one end of the bundle, where we highlight the side chain of each site in a different color.
As shown in \FIG{prefsdist}D, each of the shifted sites are more tolerant of mutations in BG505 than in BF520, indicating they are of greater structural and functional importance in BF520.
For the two shifted sites that point inward towards the bundles core (583 and 587), this importance may involve in packing.
Based on this structure, it is difficult to discern if or how the substituted site may impact the preferences at the shifted sites nearby, in part since the arginine side chain is not fully resolved.

@@

{\includegraphics[width=0.4\textwidth]{figures/shifts_on_structure/shifts_on_six_helix_bundle.pdf}}

@@

For detection of sites with faster or slower than expected evolution, we used the approach in \citet{bloom2017identification}, which is exactly modeled on the FEL approach of \citet{kosakovsky2005not} but extended to ExpCM.
This approach estimates a $P$-value that $\omega_r$ is not equal to one for each site $r$ using a likelihood-ratio test.
The actual point estimates of $\omega_r$ are unreliable for individual sites due to the limited number of observations, so we report the $P$-value that $\omega_r$ is not equal to one, which is a better indication of the strength of the statistical evidence for faster or slower than expected evolution~\citep{kosakovsky2005not,murrell2012detecting}.
For the Q-values and false discovery rate testing, we considered the tests for $\omega_r > 1$ and $\omega_r < 1$ separately.

>>>>

Specifically, we used our deep mutational scanning data to construct experimentally informed codon models (ExpCM's) for Env's evolution.
An ExpCM is a phylogenetic substitution model that incorporates the functional constraints measured in a deep mutational scanning experiment~\citep{hilton2017phydms}.
If the experiment captures much of the actual evolutionary constraint on a gene, then an ExpCM will describe the gene's natural evolution better than a standard phylogenetic codon substitution model.
The reason is that standard codon substitution models~\citep[such as the commonly used Goldman-Yang style models;][]{yang2000codon} only model functional constraint via a parameter that represents the rate of fixation of nonsynonymous protein-altering mutations relative to synonymous ones; this parameter is called dN/dS or $\omega$.
In contrast, an ExpCM accounts for the site-specific functional constraints measured in the deep mutational scan, and then additionally adds an $\omega$ parameter that now represents the relative rate of nonsynonymous to synonymous substitutions after accounting for these functional constraints~\citep{bloom2017identification,hilton2017phydms}.
Importantly, since we expect some sites in Env to be under diversifying selection from immunity, we extended the ExpCM's described in \citet{hilton2017phydms} to draw $\omega$ from a gamma distribution as is commonly done for codon-substitution models~\citep{yang2000codon}.

@@

\begin{figure}
\begin{fullwidth}
{\includegraphics[width=1.3\textwidth]{figures/BG505_prefs.pdf}}
\caption{\label{fig:BG505prefs}
Amino-acid preferences for the BG505 Env.
At each site, the height of the letter is proportional for that site's preference for that amino acid.
The top color bar indicates the region of Env (gp120 variable loop, gp120 not variable loop, or gp41).
The lower color bar indicates the evidence that the site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experiments~\citep[see the last section of the Results and][]{bloom2017identification}; we report the $P$-value for $\omega_r \ne 1$ rather than the value of $\omega_r$ itself since point estimates of these values are unreliable for individual sites due to low numbers of observations~\citep{kosakovsky2005not,murrell2012detecting}.
The letters above the logos indicate the wildtype amino acid in BG505.
Sites are numbered using the HXB2 scheme~\citep{korber1998numbering}).
This logo plot shows the site-specific amino-acid preferences for BG505 after averaging the replicates and re-scaling by the stringency parameter in \TABLE{phydms}.
The figure was generated using \texttt{dms\_tools2}~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}, which in turn utilizes \texttt{weblogo}~\citep{crooks2004weblogo}.
}
\figdata{The numerical values of the amino-acid preferences plotted in this figure are in \texttt{rescaled\_BG505\_prefs.csv.}}
\figdata{The sequence of BG505 Env and mapping from sequential (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BG505\_to\_HXB2.csv}.}
\figdata{The $\omega_r$ values and associated $P$-values for BG505 in HXB2 numbering are in \texttt{BG505\_omegabysite.tsv}.}
\end{fullwidth}
\end{figure}

@@

\figsupp[\label{figsupp:shifts_on_six_helix_bundlel}Clusters of shifted and substituted sites in context of different conformations of Env.]
{This figure gives a fine-grained view of three clusters of shifted sites in context of different conformations of Env.
In each structure, we only show side chains for shifted sites or other sites of interest, such as substituted sites.
Side chains are colored differently according to the key.
This panel shows a cluster of four shifted sites at the apex of the closed pre-fusion conformation of Env~\citep[PDB 5FYL;][]{stewart2016trimeric}.
Each of these shifted sites has also substituted.
Further, this cluster is adjacent to other substituted sites, raising the possibility of epistasis among these sites.
This region of the structure involves packing between variable loops (V1/V2 and V3) and between these loops and an adjacent Env protomer.
This packing is known to control the conformational dynamics of these loops~\cite{guttman2014cd4,powell2017plasticity}.
Notably, a recent study found that mutations at two of the shifted sites (165 and 307) resulted in Env trimers with aberrant conformations~\cite{lee2017broadly}.
Thus, it is possible that this cluster of shifted sites is involved in controlling Env's dynamics.
Indeed, although each of these shifted sites prefers hydrophobic amino acids in both homologs, one homolog often has a higher preference for bulkier aromatic amino acids than the other.
This difference could make structural sense if these residues pack in different arrangements in the different homologs, possibly due to the substitutions in this region.
However, this hypothesis is difficult to test since there is not a high-resolution structure for BF520.
This panel shows a cluster of shifted sites that are adjacent to a network of hydrophobic residues (blue/purple) that help mediate the conformational change between the closed pre-fusion state to the open CD4-bound state~\cite{ozorowski2017open}.
These sites are shown in context of both states, the latter of which is from~\citep[PDB 5VN3;][]{ozorowski2017open}.
Nearly one third of the shifted residues cluster within this structural region (9/30).
In contrast to panel (A), only a few of the shifted sites are adjacent to substitutions or have substituted themselves.
This trend is found in both the closed and open conformations, suggesting that the shifts may be primarily due to long-range epistatic interactions.
Such long-range interactions seem especially plausible in this region given that it is dynamic over a large swath or residues.
Indeed, one of the shifted sites (69) overlaps with the hydrophobic network, and mutations at another (64) have been shown to modulate the relative stability of the open and closed conformations~\cite{de2015immunogenicity}.
This panel shows a cluster of shifted sites in context of a three-helix bundle in Env, each helix of which is derived from a different monomer and which pack together to form an inter-monomer core.
These helices form the core of the post-fusion six-helix bundle shown on the right~\citep[PDB 1ENV;][]{weissenhorn1997atomic}, but are also present in a similar conformation in the closed pre-fusion state and the open CD4-bound state.
These sites are symmetric around the axis of the bundle in an arrangement that is largely similar among the different structures.
In the open conformation, one of these sites (582) is in close proximity to another site that has both shifted and substituted (520), which could account for site 582's shift in preference.
Two of the shifted sites (583 and 587) pack in the bundle's core, suggesting that the packing of residues in this region may differ between homologs.
However, unlike panel (A), where we observed a similar phenomenon, there are no substituted sites that also pack in a nearby location.
Thus, aside from site 582, it is difficult to say from a structural standpoint why these sites may have shifted.

@@

{\includegraphics[width=\textwidth]{figures/shifts_on_structure/clusters_on_structure.pdf}}

@@

For detection of sites with faster or slower than expected evolution, we used the approach in \citet{bloom2017identification}, which is exactly modeled on the FEL approach of \citet{kosakovsky2005not} but extended to ExpCM.
This approach estimates a $P$-value that $\omega_r$ is not equal to one for each site $r$ using a likelihood-ratio test.
For the Q-values and false discovery rate testing, we considered the tests for $\omega_r > 1$ and $\omega_r < 1$ separately.
